Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-03-10 paper/rebuttal/rebuttal.tex

\addbibresource{rebuttal.bib}

%
% HEADER
%

\newcounter{comment}[subsection]

%
% COLORBOX
%
\usepackage{tcolorbox}
\tcbuselibrary{breakable}
\newtcolorbox[]{sectionbox}[1]{colback=white,
 				       												 colframe=black,
 				       												 colbacktitle=white!50!gray,
 				       												 coltitle=black,
 				       												 fonttitle=\bfseries,
 				       												 phantom=\refstepcounter{comment},
 				       												 title={Comment \arabic{section}.\arabic{comment}},
 				       												 list entry={Comment \arabic{section}.\arabic{comment}\quad}, breakable}

### 2021-03-10 paper/rebuttal/rebuttal.tex

Kind Regards,

Dakota, Jisung, Sadamori, Rodrigo, Woo-Sung, Staša, and YY

\pagenumbering{arabic}
\clearpage
\newgeometry{margin=1in}
\setcounter{page}{1}
\tableofcontents
\clearpage

\section{Theoretical justification of \textit{word2vec} and the Gravity Model}
\label{sec:theory}
\thispagestyle{empty}
Why does the \textit{word2vec} with skip-gram and negative sampling (SGNS) model work so well to model mobility?
Here, we demonstrate that SGNS and the gravity model of mobility are mathematically equivalent.
This connection provides strong theoretical justification to our choice of \textit{word2vec} to model mobility, and constitutes a major breakthrough of our understanding of neural embeddings.
In addition, this motivates the use of SGNS for the future study of human mobility.

\subsection{\textit{word2vec} algorithm}
Imagine a trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), where $a_{t}$ is the $t$\textsuperscript{th} location in the trajectory. A location, $a_{t}$, is considered to have context locations, $a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}$, that appear in the window surrounding $a_t$ up to a time lag of $w$, where $w$ is the window size parameter truncated at $t - w \geq 0$ and $t + w \leq T$.
Then, the skip-gram \textit{word2vec} model learns probabilities $p(a_{t + \tau} \vert a_{t})$, where $-w\leq \tau\leq w$ and $\tau \neq 0$,  by maximizing its log likelihood given by
%
% MAIN W2V Equation
%
\begin{equation}
	{\cal J} = \frac{1}{T}\sum_{t = 1}^{T} \sum_{-w \leq \tau \leq w, \tau \neq 0} \log P(a_{t + \tau} \vert a_{t}), \label{eq:word2vec-obj}
\end{equation}
where,

%
% CONDITIONAL PROBABILITY EQ
%
\begin{equation}
	P(j \given i) = \frac{\exp(\vect{u}_j \cdot \vect{v}_{i})}{Z_i}, \label{eq:cond_prob_w2v}
\end{equation}
where $\vect{v}$ and $\vect{u}$ are the ``in-vector'' and ``out-vector'', respectively,  $Z_i=\sum_{j' \in \mathcal{A}} \exp(\vect{u}_{j'} \cdot \vect{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations. In general, calculating $Z_i$ is computationally expensive, so two kinds of approximations are commonly used:
hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

@@

\subsection{An implicit bias in the negative sampling}

Although negative sampling is the most common approximation, it is a biased estimator~\autocite{Chia2010,Dyer2014} that does not yield the optimal embeddings in terms of the objective function, ${\cal J}$.
This bias in negative sampling turns out to be key to the connection between word2vec and the gravity model.

@@

\rcomment{%
	The methodology has potential limitations in terms of reproducibility and robustness, which should be addressed. This is no specific fault of the authors, but both the SGNS/word2vec algorithm and the UMAP algorithm are inherently stochastic. This means that these approaches involve inherent randomness, and for researchers to replicate the exact effect seen in this manuscript, one would need to ensure the exact same random seeds etc. were used. In my view, this is a serious limitation, especially given that the UMAP results and learned embeddings are reified and treated as first-class objects of scientific study.
}

\response{Thank you for this comment. While robustness is a general concern for these methods, we found that their application to our datasets were in fact quite robust, and observed no evidence that findings were the result of methodological artifacts. This is confirmed in extra work done in response to these reviews.

	We re-ran all of our results using different starting seeds and found all results were qualitatively similar, and almost quantitatively identical.
	In addition, we expanded the set of experimental parameters, training 18 different embeddings with different dimensions, window sizes, and now, $\gamma$.
	We found that while performance varies across each combination of parameters, the main results remain qualitatively similar.
	While we can run more combinations of parameters, and more iterations of \textit{word2vec}, we see no evidence that results will drastically vary.

	We believe that the ratio of the number of organizations represented in our dataset to the total number of mobility trajectories makes our results especially robust.
	About 8,661 organizations are represented across over three million career trajectories of scientists, such that nearly every organization appears enough times, and in enough contexts, to produce a robust vector representation.
	However, this may not hold true for all problem domains—for example, natural language includes many millions of tokens, many of which will occur only a few times, allowing stochasticity to have a much stronger effect on the final representation.
	Heeding this as a general concern, we have added the following sentence to our limitations paragraph in our conclusion:

@@

\rcomment{%
	Although the paper is well written and explores a very interesting problem, I have certain criticism for the used methodology. After obtaining a representation of each dataset, the authors defined an embedding distance as the cosine distance between vectors in the space, while defining distance as the geographical distance in the original, spatially embedded datasets. Subsequently they used the Gravity Law to estimate the flux between location pairs of vector pairs in the spatial and embedded space respectively. The Gravity Law, applied on mobility, assumes that the number of people moving between two populations is proportional to the multiplicative of the sizes of the populations and a deterrence function. Research during the last decades on transportation, spatial networks and human mobility suggests that this deterrence function may take a functional form of an inverse exponential or an inverse power-law, as it has been acknowledged by the authors. However, the emergence of these different deterrence functions can be due to fundamentally different mechanisms (for a review on the topic see Barthélemy, Physics Reports 499, 1, 2011, a paper not cited in the manuscript). For example, the different cost functions (linear or logarithmic) to create link at a given distance, the spatial granularity of the observation, or even the heterogeneity of the moving individuals in terms of various factors, just to mention a few examples. Instead, the authors of the actual paper arbitrarily assumed a power-law deterrence function in the spatial case and an exponential deterrence function in the embedding space without further arguing their choice (other than fitting on the data). Furthermore, assuming different deterrence functions for the flux estimation using Gravity Law makes the estimated fluxes incomparable, resulting the invalidation of results depicted in Figure 2.
}

\response{Thank you for noting this.
	We carefully read out the Barthélemy, Physics Reports 499, 1, 2011, and found it incredibly educational, referencing it in our manuscript.		
	It has been argued that the power-law decay function is appropriate for modelling geographic distance the model is more suitable for analyzing large, complex, and scale-free spatial systems with long-range spatial memory, whereas the exponential gravity model fits well with the simpler system with no spatial memory~\autocite{chen2015distance}.
	We report the performance of the flux estimation and prediction using both deterrence functions (see Section 2, above), finding that although the absolute values vary, the qualitative pattern remains the same---the embedding distance outperforms geographic distance regardless of the function used.
	 We also elaborate on the connection between the embedding distance and exponential model during out discussion on \textit{word2vec} and the gravity model of mobility (see Section 1, above).
	 In addition, we add the following line to our manuscript, noting these considerations,

@@

	What is notable about our findings, therefore, is that even though a model trained with $w = 1$ generates fewer training samples, it still outperforms settings with $w > 1$, which use more training samples, and which should capture higher-order structures in the data.
	We speculate that this is because the word2vec model with $w = 1$ is equivalent to the gravity model of mobility (see Section 1).
}

\rcomment{
	The other very important parameter of the SkipGram method is the embedding dimension, which is explained in the Methods but it is not explored. E.g. to choose the ideal dimension for embedding one may sweep through a broad range of possible dimensions and not only explore three values as described in the SI (Fig.S4).
	Their main observations in section “Embeddings provide functional distance between locations” is that “…our results demonstrate that, consistently, the embedding distance better captures patterns of actual mobility than does the geographic distance.”. This is a great verification of their method but it is somewhat expected from an embedding, which is built on mobility trajectories between places, as compared to cases based on simply geographic distance.

}
\response{
	Thank you for these comments.
	First, we added a line to the methods section detailing the \textit{word2vec} we used, and noting that these achieved the best performance: \say{\textit{For main results regarding scientific mobility, we used $d=300$ and $w=1$, which were the parameters that best explained the flux between locations, though results were robust across different settings (Fig....)}}.

	A full search of the hyper-parameter space would certainly be interesting and useful, however we believe that set of hyper-parameters we considered here is sufficient to establish the utility of the approach, and demonstrate that performance is only slightly different based on different hyper-parameters.
	There are previous studies on the optimal dimensionality of word embedding \autocite{yin2018understand, yin2018dimensionality}. Also,
	we note that in previous work from our group, we explored the optimal dimensions for graph embedding~\autocite{gu2020dimensions}, finding that while a sufficient number of dimensions is necessary to capture the structure, increasing the dimensions does not heavily deteriorate the performance, a property common across deep learning methods.
	Given that there is little performance difference between $d = 100$ and $d = 300$, we opt to use the larger embeddings at $d = 300$, which are large enough to capture the complexity of mobility, yet small enough to allow an efficient computational workflow.
}

@@

\response{We have updated the relevant line of the caption to read as follows:
	“The gravity model with the exponential decay function slightly outperforms that with a power-decay function except in the case of international-only mobility, for which power-decay performs slightly better.”
}

%
%
% Response to reviewer 3
\clearpage
\setcounter{section}{3}
\subsection{Response to Referee 3}
\setcounter{section}{3}

\rcomment{The analysis does not compare embeddings, which would not be a problem if they were making
	an inference other than we can embed micro-mobility to trace macro-mobility. They use an
	efficient (Google engineers developed it), single-layer embedding from 2013, which produces
	linear associations. But the paper doesn’t take advantage of the revolution in deep neural
	network models that now produce powerful, nonlinear and noncontiguous (e.g., using the
	attention architecture) models that now dominate analysis of language, and are making very
	power inroads in machine vision. To be clear, deeper models require much data and computational power to tune—they are an engineering feat—but if developed in these settings, they would likely dramatically improve the prediction of flux. They take into account directionality (e.g., a person moving from one opportunity to another) and would independently capture people how people distinctly move up and down the hierarchy or from place to place (e.g., a in bi-directional model like BERT). A lack of discussion or demonstration of which model works best--and the unfounded inference that direction does not matter felt weak without a demonstration that they would not have worked better. They did and do work much better for language models and for image generation...why would they not work for analysis of human mobility? (I’m sure they would, but this would time and independent development.)
}

\response{Thank you for this insightful comment!
	We think that there is a great deal of potential in methods like BERT for representing mobility data.
	What initially drew us to \textit{word2vec}, however, was its simplicity, ubiquity, and mathematical tractability (at least compared to other, more sophisticated approaches).
	This last attribute proved very useful, as in exploring the mathematical properties of the model, we discovered a deep connection between \textit{word2vec} and the gravity model of mobility, which explains why it learns gravity-like relationships between entities (see Section 1, this letter).
	We believe that this new discovery, now included in the manuscript, strongly grounds our use of \textit{word2vec} here.

### 2021-03-10 paper/current/main.tex
update main paper
tweaks in main text

	Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
	Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
	However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
	As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity is of paramount importance.
	Here, we illustrate a mathematical connection between the neural embedding model \textit{word2ec}~\autocite{mikolov2013word2vec} and the gravity model of mobility and, using three distinct human trajectory datasets, demonstrate that a neural embedding model can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility.
	Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncover cultural and linguistic relations, and even academic prestige, at multiple levels of granularity.
	Furthermore, the embedding vectors reveal universal relationships between organizational characteristics and their place in the global landscape of scientific mobility.
	The ability to learn scalable, dense, and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

Although negative sampling is the most common approximation, it is a biased estimator~\autocite{Chia2010,Dyer2014} and fits a different probability model. When taking into account this bias, {\it word2vec} with the skip-gram and negative sampling fits a probability model given by
\begin{align}
	P\left(j \given i \right):= \frac{P^ \gamma(j) \exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}, \label{eq:prob_w2v_ng0}
\end{align}
where we redefine the normalization constant as $Z'_i=\sum_{j' \in \mathcal{A}} P^ \gamma (j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$. (See the Methods and Supporting Information for the full derivation).

@@

In practice, because we only have limited amounts of noisy data and the optimization may not find the true optimum, the mathematical result may only approximately hold.
Indeed, we find that the in- and out-vectors tend to be different and that the cosine similarity tends to better capture real-world mobility than the dot similarity.
This result echos other applications of word embedding, such as word analogy testing~\autocite{levy2015improving}, in which cosine distance also outperformed dot similarity.
Nevertheless, a model with dot similarity is still the second-best after cosine similarity (Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}), and the embedding distance still outperforms all alternatives we considered.

@@

%
% Acknowledgements
%
\subsection*{Acknowledgments}
We thank the Center for Science and Technology Studies at Leiden University for managing and making available the dataset of scientific mobility. We also thank the  Goodchoice Company LTD. for making available the dataset of Korean accommodation reservation data.
For their comments, we thank Guillaume Cabanac, Cassidy R. Sugimoto, Vincent Lariviére, Alessandro Flammini, Filippo Menczer, Lili Miao, Xiaoran Yan, and Inho Hong.
This material is based upon work supported by the Air Force Office of Scientific Research under award number FA9550-19-1-0391.
Rodrigo Costas is partially funded by the South African DST-NRF Centre of Excellence in Scientometrics and Science, Technology and Innovation Policy (SciSTIP).

>>>>

	Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
	Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
	However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
	As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity is of paramount importance.
	Here, we reveal a mathematical connection between the neural embedding model \textit{word2ec}~\autocite{mikolov2013word2vec} and the gravity model of mobility and, using three distinct human trajectory datasets, demonstrate that a neural embedding model can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility.
	Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncover cultural and linguistic relations, and even academic prestige, at multiple levels of granularity.
	Furthermore, the embedding vectors reveal universal relationships between organizational characteristics and their place in the global landscape of scientific mobility.
	The ability to learn scalable, dense, and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

Although negative sampling is the most common approximation, it is a biased estimator~\autocite{Chia2010,Dyer2014} and fits a different probability model. When taking into account this bias, {\it word2vec} with skip-gram and negative sampling fits a probability model given by
\begin{align}
	P\left(j \given i \right):= \frac{P^ \gamma(j) \exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}, \label{eq:prob_w2v_ng0}
\end{align}
where we redefine the normalization constant as $Z'_i=\sum_{j' \in \mathcal{A}} P^ \gamma (j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$. (See the Methods and Supporting Information for the full derivation).

@@

In practice, because we only have limited amounts of noisy data and the optimization may not find the true optimum, the mathematical result may only approximately hold.
Indeed, we find that the in- and out-vectors tend to be different and that the cosine similarity tends to better capture real-world mobility than the dot similarity.
This result echos other applications of word embedding, such as word analogy testing~\autocite{levy2015improving}, in which cosine distance also outperformed dot similarity.
Nevertheless, a model with dot similarity has the second-best performance after cosine similarity (Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}), and the embedding distance still outperforms all alternatives we considered.

@@

%
% Acknowledgements
%
\subsection*{Acknowledgments}
We thank the Center for Science and Technology Studies at Leiden University for managing and making available the dataset of scientific mobility. We also thank the  Goodchoice Company LTD. for making available the dataset of Korean accommodation reservation data.
For their comments, we thank Guillaume Cabanac, Cassidy R. Sugimoto, Vincent Lariviére, Alessandro Flammini, Filippo Menczer, Lili Miao, Xiaoran Yan, Inho Hong, and Esteban Moro Egido.
This material is based upon work supported by the Air Force Office of Scientific Research under award number FA9550-19-1-0391.
Rodrigo Costas is partially funded by the South African DST-NRF Centre of Excellence in Scientometrics and Science, Technology and Innovation Policy (SciSTIP).
